---
title: "The buzz on using bHIVE"
author: 
  name: Nick Borcherding
  email: ncborch@gmail.com
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
package: bHIVE
vignette: >
  %\VignetteIndexEntry{The buzz on using bHIVE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
library(BiocStyle)
library(bHIVE)
library(ggplot2)
library(viridis)
library(caret)
```

# Introduction

The **bHIVE** package implements an Artificial Immune Network (AI-Net) algorithm for clustering, classification, and regression tasks. Inspired by biological immune systems, bHIVE uses principles like clonal selection, mutation, and suppression to analyze and model data. 

This vignette demonstrates how to:
1. Perform clustering, classification, and regression using `bHIVE`.
2. Tune hyperparameters using `swarmbHIVE`.
3. Use the `caret` wrapper for easy integration with machine learning workflows.
4. Use multilayered immune networks with `honeycombHIVE`
4. Visualize results with `ggplot2`.

## Parametes for `bHIVE()`

The behavior of the `bHIVE` function can be fine-tuned using a range of hyperparameters. Below is a description of the key parameters:

| **Parameter**         | **Description**                                                                                               |
|------------------------|-------------------------------------------------------------------------------------------------------------|
| `X`                   | A numeric matrix or data frame of input features (rows are observations, columns are features).              |
| `y`                   | (Optional) Target vector for classification (factor) or regression (numeric). If `NULL`, clustering is performed. |
| `task`                | Specifies the task: `"clustering"`, `"classification"`, or `"regression"`.                                   |
| `nAntibodies`         | Number of initial antibodies in the population. Larger values increase diversity but add computational cost.  |
| `beta`                | Clone multiplier. Determines how many clones are generated for top-matching antibodies.                      |
| `epsilon`             | Suppression threshold. Antibodies closer than `epsilon` are considered redundant and removed.                |
| `maxIter`             | Maximum number of iterations for the algorithm.                                                             |
| `affinityFunc`        | Affinity (similarity) function. Options include `"gaussian"`, `"laplace"`, `"polynomial"`, `"cosine"`, `"hamming"`. |
| `distFunc`            | Distance function for clustering and suppression. Options include `"euclidean"`, `"manhattan"`, `"cosine"`, `"minkowski"`, `"hamming"`. |
| `affinityParams`      | A list of optional parameters for the affinity/distance functions.                                           |
| `mutationDecay`       | Factor controlling how the mutation rate decays over iterations. Default is `1.0` (no decay).                |
| `mutationMin`         | Minimum mutation rate to avoid zero mutation.                                                               |
| `maxClones`           | Maximum number of clones per antibody. Default is unlimited (`Inf`).                                         |
| `stopTolerance`       | Tolerance for stopping the algorithm if the antibody population size stabilizes.                             |
| `noImprovementLimit`  | Number of iterations without improvement before early stopping.                                              |
| `initMethod`          | Method for initializing antibodies. Options: `"sample"` (randomly selects rows from `X`) or `"random"` (Gaussian noise). |
| `k`                   | Number of top-matching antibodies to consider during cloning.                                               |
| `seed`                | Random seed for reproducibility.                                                                             |
| `verbose`             | Logical. If `TRUE`, prints progress messages for each iteration.                                             |

---

### How `bHIVE` Works

1. **Affinity Calculation**: Measures the similarity between each data point and antibodies using a specified `affinityFunc`.
2. **Clonal Expansion and Mutation**: Generates clones of top-matching antibodies, introducing diversity via mutation.
3. **Network Suppression**: Removes antibodies that are too similar to maintain diversity in the population.
4. **Assignment**: Assigns data points to antibodies (clusters, labels, or numeric predictions).

### Example of Parameter Selection

The following example demonstrates how to configure the `bHIVE` function for clustering, emphasizing the impact of key parameters:

```{r tidy = FALSE, eval=FALSE}
# Load the Iris dataset
data(iris)
X <- as.matrix(iris[, 1:4])

# Configure bHIVE parameters for clustering
set.seed(42)
res <- bHIVE(
  X = X,                      # Input data
  task = "clustering",        # Task type
  nAntibodies = 20,           # Number of antibodies
  beta = 5,                   # Clone multiplier
  epsilon = 0.01,             # Suppression threshold
  maxIter = 20,               # Maximum iterations
  affinityFunc = "gaussian",  # Affinity function
  distFunc = "euclidean",     # Distance function
  verbose = TRUE              # Print progress
)
```

# Applications

## Clustering 

**Clustering** is an unsupervised learning task where we group similar data points based on their features. In this example, we cluster the numeric features of the Iris dataset using `bHIVE`. 

```{r tidy = FALSE}
# Load Iris dataset
data(iris)
X <- as.matrix(iris[, 1:4])

# Perform clustering
set.seed(42)
res <- bHIVE(X = X, 
             task = "clustering", 
             nAntibodies = 10, 
             beta = 5, 
             epsilon = 0.05, 
             maxIter = 20, 
             k = 3,
             verbose = FALSE)

# Add cluster assignments to the data
iris$Cluster <- as.factor(res$assignments)

# Visualize clusters
ggplot(iris, aes(x = Sepal.Length, 
                 y = Sepal.Width, 
                 color = Cluster)) +
    geom_point(size = 3) +
    labs(title = "Clustering Results with bHIVE", 
         x = "Sepal Length", 
         y = "Sepal Width") +
    scale_color_viridis(discrete = TRUE)
    theme_minimal()
```

## Classification

**Classification** is a supervised learning task where data points are assigned to predefined categories based on their features. Here, we classify the species of Iris flowers using `bHIVE`.

```{r tidy = FALSE}
# Classification setup
y <- iris$Species

# Perform classification
set.seed(42)
res <- bHIVE(X = X, 
             y = y, 
             task = "classification", 
             nAntibodies = 50, 
             beta = 4, 
             epsilon = 0.02, 
             maxIter = 64, 
             initMethod = "random",
             k = 4,
             verbose = FALSE)

# Compare predicted vs actual
table(Predicted = res$assignments, Actual = y)

# Visualize classification results
iris$Predicted <- res$assignments
ggplot(iris, aes(x = Sepal.Length, 
                 y = Sepal.Width, 
                 color = Predicted, 
                 shape = Species)) +
    geom_point(size = 3) +
    labs(title = "Classification Results with bHIVE", 
         x = "Sepal Length", 
         y = "Sepal Width") +
    scale_color_viridis(discrete = TRUE)
    theme_minimal()
```

## Regression

**Regression** is a supervised learning task where the goal is to predict continuous numeric outcomes. In this example, we use the Boston housing dataset to predict the median home value (medv) using `bHIVE`.

```{r tidy = FALSE}
if (!requireNamespace("MASS", quietly = TRUE)) install.packages("MASS")
library(MASS)
data(Boston)
X <- as.matrix(Boston[, -14])
y <- Boston$medv

# Perform regression
set.seed(42)
res <- bHIVE(X = X, 
             y = y, 
             task = "regression", 
             nAntibodies = 50, 
             beta = 5, 
             epsilon = 0.01, 
             maxIter = 100, 
             initMethod = "random",
             k = 3,
             verbose = FALSE)

# Compare predicted vs actual
cor(res$assignments, y)

# Visualize regression results
Boston$Predicted <- res$assignments
ggplot(Boston, aes(x = y, y = Predicted)) +
    geom_point() +
    geom_smooth(method = "lm", color = "blue", se = FALSE) +
    labs(title = "Regression Results with bHIVE", 
         x = "Actual Values (medv)", 
         y = "Predicted Values") +
    theme_minimal()

```

## Hyperparameter Tuning

Tuning hyperparameters is crucial for optimizing the performance of machine learning algorithms. In this example, we perform hyperparameter tuning for clustering using `swarmbHIVE`.

```{r tidy = FALSE}
grid <- expand.grid(
    nAntibodies = c(10, 20),
    beta = c(3, 5),
    epsilon = c(0.01, 0.05)
)

# Perform tuning
set.seed(42)
tuning_results <- swarmbHIVE(X = X, 
                             task = "clustering", 
                             grid = grid, 
                             metric = "silhouette",
                             verbose = FALSE)

# Best parameters
tuning_results$best_params

# Visualize tuning results
ggplot(tuning_results$results, aes(x = nAntibodies, y = metric_value, color = factor(beta))) +
    geom_line() +
    geom_point(size = 3) +
    labs(title = "Hyperparameter Tuning Results", 
         x = "Number of Antibodies", 
         y = "Silhouette Score", 
         color = "Beta") +
    scale_color_viridis(discrete = TRUE) + 
    theme_minimal()
```

## Using the `caret` wrapper 

The `bHIVE` package provides a `caret` wrapper for seamless integration with the `caret` framework, allowing for easy cross-validation and hyperparameter tuning. Here, we demonstrate regression on the Boston housing dataset.

```{r tidy = FALSE}
# Train using caret
train_control <- trainControl(method = "cv", number = 2)
set.seed(42)
model <- train(x = X,
               y = y,
               method = bHIVEmodel,
               trControl = train_control,
               tuneGrid = expand.grid(
                    nAntibodies = c(10, 20),
                    beta = c(3, 5),
                    epsilon = c(0.01, 0.05)),
                verbose = FALSE
)

# Visualize caret results
ggplot(model) +
    labs(title = "Caret Tuning Results for bHIVE", 
         x = "Hyperparameter Combination", 
         y = "Performance Metric") +
    scale_color_viridis(discrete = TRUE)
    theme_minimal()
```
## honeycombHIVE: Multilayered bHIVVES

In `honeycombHIVE`, clustering proceeds hierarchically across multiple layers:

* Each layer generates a refined set of cluster centroids based on the results of the previous layer.  
* At layer $L$, the centroids $A_L$ are derived from the means of data points assigned to each cluster at layer $L−1$.  

```{r tidy=FALSE}
# Load the Iris dataset
data(iris)
X <- as.matrix(iris[, 1:4])

# Run honeycombHIVE for clustering
res <- honeycombHIVE(X = X, 
                     task = "clustering", 
                     epsilon = 0.05,
                     layers = 3, 
                     nAntibodies = 30, 
                     beta = 5, 
                     maxIter = 10, 
                     verbose = FALSE)

# Visualize results from each layer
for (i in seq_along(res)) {
  
  # Create a data frame for plotting; add original Sepal.Length and Sepal.Width
  plot_df <- data.frame(
    Sepal.Length = iris$Sepal.Length, 
    Sepal.Width  = iris$Sepal.Width,
    Cluster      = factor(res[[i]]$membership)  # cluster labels from layer i
  )
  
  # Basic ggplot scatter plot
  plot <- ggplot(plot_df, aes(x = Sepal.Length, 
                              y = Sepal.Width, 
                              color = Cluster)) +
    geom_point(size = 3) +
    labs(
      title = paste("honeycombHIVE Clustering - Layer", i),
      x     = "Sepal Length",
      y     = "Sepal Width"
    ) +
    theme_minimal() +
    scale_color_viridis(discrete = TRUE)
  
  print(plot)
}
```

# Conclusions

The `bHIVE` package is a versatile tool for clustering, classification, and regression tasks. Its integration with caret simplifies hyperparameter tuning and cross-validation, making it suitable for a variety of datasets and use cases. If you have any questions, comments, or suggestions, please visit the [GitHub repository](https://github.com/ncborcherding/bHIVE).

```{r}
sessionInfo()
```
