---
title: "Advanced Tuning & Workflows"
author:
  name: Nick Borcherding
  email: ncborch@gmail.com
  affiliation: Washington University in St. Louis, School of Medicine, St. Louis, MO, USA
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc_float: true
---

```{r, echo=FALSE, results="hide", message=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE, tidy = FALSE)
library(bHIVE)
library(ggplot2)
library(viridis)
```

# Introduction

This vignette covers advanced usage patterns for bHIVE:

1. Hyperparameter tuning with `swarmbHIVE()` across tasks
2. Multilayer hierarchical refinement with `honeycombHIVE()`
3. Gradient-based prototype refinement with `refineB()`
4. Integration with the `caret` framework
5. Visualization with `visualizeHIVE()`

# Hyperparameter Tuning with swarmbHIVE

`swarmbHIVE()` performs grid search over bHIVE hyperparameters and evaluates
each combination using a task-specific metric.

## Available Metrics

| Task | Metrics | Direction |
|:-----|:--------|:----------|
| Clustering | `"silhouette"`, `"davies_bouldin"`, `"calinski_harabasz"` | Higher is better (except D-B) |
| Classification | `"accuracy"`, `"balanced_accuracy"`, `"f1"`, `"kappa"` | Higher is better |

## Clustering Tuning

```{r swarm-clustering, fig.width=7, fig.height=4}
data(iris)
X <- as.matrix(iris[, 1:4])

grid <- expand.grid(
  nAntibodies = c(10, 20, 30),
  beta        = c(3, 5),
  epsilon     = c(0.01, 0.05)
)

set.seed(42)
tuning <- swarmbHIVE(X = X, task = "clustering", grid = grid,
                     metric = "silhouette", maxIter = 15, verbose = FALSE)

ggplot(tuning$results, aes(x = nAntibodies, y = metric_value,
                            color = factor(beta))) +
  geom_line() +
  geom_point(aes(shape = factor(epsilon)), size = 3) +
  labs(title = "Clustering: Silhouette Score",
       x = "Number of Antibodies", y = "Silhouette",
       color = "Beta", shape = "Epsilon") +
  scale_color_viridis(discrete = TRUE) +
  theme_minimal()
```

```{r swarm-best}
tuning$best_params
```

## Classification Tuning

```{r swarm-classification}
y <- iris$Species
grid_cls <- expand.grid(
  nAntibodies = c(15, 30, 50),
  beta        = c(3, 5, 10),
  epsilon     = c(0.01, 0.05)
)

set.seed(42)
tuning_cls <- swarmbHIVE(X = X, y = y, task = "classification",
                         grid = grid_cls, metric = "accuracy",
                         maxIter = 15, verbose = FALSE)
tuning_cls$best_params
```

## Parallelization with BiocParallel

`swarmbHIVE()` accepts a `BPPARAM` argument for parallel execution. This
is especially useful when the grid is large:

```{r parallel-example, eval=FALSE}
library(BiocParallel)

# Use 4 cores
tuning <- swarmbHIVE(
  X = X, y = y, task = "classification",
  grid = grid_cls, metric = "accuracy",
  BPPARAM = MulticoreParam(4),
  verbose = FALSE
)
```

# Multilayer Refinement with honeycombHIVE

`honeycombHIVE()` runs bHIVE iteratively across multiple layers. Each layer
produces prototypes that become the input for the next layer, creating a
hierarchical compression of the data.

## Basic Multilayer Clustering

```{r honeycomb-clustering, fig.width=8, fig.height=4}
set.seed(42)
res <- honeycombHIVE(X = X, task = "clustering",
                     layers = 3, nAntibodies = 30,
                     beta = 5, epsilon = 0.05,
                     maxIter = 10, verbose = FALSE)

# Each layer progressively refines cluster structure
layer_info <- sapply(res, function(r) {
  c(n_prototypes = nrow(r$antibodies),
    n_clusters = length(unique(r$membership)))
})
layer_info
```

## Collapse Methods

The `collapseMethod` parameter controls how clusters are summarized into
prototypes for the next layer:

| Method | Description |
|:-------|:------------|
| `"centroid"` | Mean of cluster members (default) |
| `"medoid"` | Actual data point closest to center |
| `"median"` | Coordinate-wise median |

```{r collapse-comparison}
set.seed(42)
res_centroid <- honeycombHIVE(X = X, task = "clustering", layers = 2,
                              nAntibodies = 20, collapseMethod = "centroid",
                              verbose = FALSE)
res_medoid <- honeycombHIVE(X = X, task = "clustering", layers = 2,
                            nAntibodies = 20, collapseMethod = "medoid",
                            verbose = FALSE)

cat("Centroid clusters:", length(unique(res_centroid[[2]]$membership)), "\n")
cat("Medoid clusters:  ", length(unique(res_medoid[[2]]$membership)), "\n")
```

## With Gradient Refinement

Setting `refine = TRUE` applies gradient-based updates via `refineB()`
after each layer's bHIVE pass. This fine-tunes prototype positions
before collapsing.

```{r honeycomb-refine}
data(iris)
X_iris <- as.matrix(iris[, 1:4])
y_iris <- iris$Species

set.seed(42)
res_plain <- honeycombHIVE(X = X_iris, y = y_iris, task = "classification",
                           layers = 3, nAntibodies = 30, verbose = FALSE)

res_refined <- honeycombHIVE(X = X_iris, y = y_iris, task = "classification",
                             layers = 3, nAntibodies = 30,
                             refine = TRUE, refineOptimizer = "adam",
                             refineLoss = "categorical_crossentropy",
                             refineSteps = 5, refineLR = 0.01, verbose = FALSE)

acc_plain   <- mean(res_plain[[3]]$predictions   == as.character(y_iris))
acc_refined <- mean(res_refined[[3]]$predictions == as.character(y_iris))
cat("Accuracy (plain):  ", round(acc_plain, 3), "\n")
cat("Accuracy (refined):", round(acc_refined, 3), "\n")
```

# Gradient Refinement with refineB

`refineB()` takes trained antibody positions and fine-tunes them using
gradient-based optimization. It supports 5 optimizers and 8 loss functions.

## Optimizers

| Optimizer | Key Parameter | Notes |
|:----------|:--------------|:------|
| `"sgd"` | `lr` | Simple, good baseline |
| `"momentum"` | `momentum_coef` | Accelerates SGD |
| `"adagrad"` | -- | Per-parameter adaptive learning rates |
| `"adam"` | `beta1`, `beta2` | Best general-purpose optimizer |
| `"rmsprop"` | `rmsprop_decay` | Good for non-stationary objectives |

## Loss Functions

| Loss | Tasks | Description |
|:-----|:------|:------------|
| `"mae"` | All | Mean absolute error (sign-based pull/push) |
| `"categorical_crossentropy"` | Classification | Cross-entropy loss |
| `"binary_crossentropy"` | Classification | Binary cross-entropy |
| `"kullback_leibler"` | Classification | KL divergence |
| `"cosine"` | Classification | Cosine distance loss |

## Comparing Optimizers

```{r refine-comparison, fig.width=8, fig.height=4}
X <- as.matrix(iris[, 1:4])
y <- iris$Species

set.seed(42)
res <- bHIVE(X, y, task = "classification", nAntibodies = 10,
             beta = 5, epsilon = 0.05, initMethod = "random",
             k = 4, verbose = FALSE)

Ab <- res$antibodies
colnames(Ab) <- colnames(X)
assignments <- as.integer(factor(res$assignments,
                                 levels = unique(res$assignments)))

pca <- prcomp(X, scale. = TRUE)
X_pca <- pca$x[, 1:2]
A_orig_pca <- predict(pca, Ab)

optimizers <- c("sgd", "adam", "rmsprop")
refined_list <- lapply(optimizers, function(opt) {
  Ab_refined <- refineB(A = Ab, X = X, y = y,
                        assignments = assignments,
                        task = "classification",
                        loss = "categorical_crossentropy",
                        optimizer = opt, steps = 5, lr = 0.01,
                        verbose = FALSE)
  colnames(Ab_refined) <- colnames(X)
  A_ref_pca <- predict(pca, Ab_refined)
  data.frame(optimizer = opt,
             PC1_after = A_ref_pca[, 1],
             PC2_after = A_ref_pca[, 2])
})

refined_df <- do.call(rbind, refined_list)
refined_df$ID <- factor(rep(seq_len(nrow(Ab)), times = length(optimizers)))

proto_df <- data.frame(ID = factor(seq_len(nrow(A_orig_pca))),
                       PC1_before = A_orig_pca[, 1],
                       PC2_before = A_orig_pca[, 2])

merged <- merge(proto_df, refined_df, by = "ID")

ggplot() +
  geom_point(data = data.frame(PC1 = X_pca[, 1], PC2 = X_pca[, 2],
                                Cluster = factor(assignments)),
             aes(x = PC1, y = PC2, color = Cluster), alpha = 0.4) +
  geom_point(data = proto_df,
             aes(x = PC1_before, y = PC2_before),
             color = "#440154", size = 4) +
  geom_point(data = merged,
             aes(x = PC1_after, y = PC2_after),
             color = "#FDE725", size = 4) +
  geom_segment(data = merged,
               aes(x = PC1_before, y = PC2_before,
                   xend = PC1_after, yend = PC2_after),
               arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
               color = "black", linewidth = 0.5) +
  scale_color_viridis(discrete = TRUE) +
  labs(title = "Prototype Movement by Optimizer",
       subtitle = "Purple = before, Yellow = after refinement",
       x = "PC1", y = "PC2") +
  facet_wrap(~optimizer, nrow = 1,
             labeller = as_labeller(c(sgd = "SGD", adam = "Adam",
                                      rmsprop = "RMSProp"))) +
  theme_minimal()
```

# caret Integration

bHIVE provides two caret-compatible model objects for use with `train()`:

- `bHIVEmodel` -- single-pass bHIVE with tuning over `nAntibodies`,
  `beta`, and `epsilon`
- `honeycombHIVEmodel` -- multilayer bHIVE with additional tuning over
  `layers`, `refineOptimizer`, `refineSteps`, and `refineLR`

## Basic caret Workflow

```{r caret-basic, fig.width=7, fig.height=5}
library(caret)

data(iris)
X_iris <- as.matrix(iris[, 1:4])
y_iris <- iris$Species

set.seed(42)
idx <- sample(nrow(X_iris), nrow(X_iris) * 0.7)
X_train <- X_iris[idx, ]
X_test  <- X_iris[-idx, ]
y_train <- y_iris[idx]
y_test  <- y_iris[-idx]

ctrl <- trainControl(method = "cv", number = 3)

model <- train(
  x = X_train, y = y_train,
  method = bHIVEmodel,
  trControl = ctrl,
  tuneGrid = expand.grid(
    nAntibodies = c(10, 20),
    beta        = c(3, 5),
    epsilon     = c(0.01, 0.05)
  ),
  verbose = FALSE
)

ggplot(model) +
  labs(title = "caret Cross-Validation Results") +
  scale_color_viridis(discrete = TRUE) +
  theme_minimal()
```

## Prediction on Held-Out Data

```{r caret-predict}
preds <- predict(model, newdata = X_test)
table(Predicted = preds, Actual = y_test)
```

## honeycombHIVE caret Model

The multilayer caret model adds layer count and refinement parameters to
the tuning grid:

```{r caret-honeycomb, eval=FALSE}
model_hc <- train(
  x = X_train, y = y_train,
  method = honeycombHIVEmodel,
  trControl = trainControl(method = "cv", number = 3),
  tuneGrid = expand.grid(
    nAntibodies = c(20, 30),
    beta        = 5,
    epsilon     = 0.05,
    layers      = c(2, 3),
    refineOptimizer = "adam",
    refineSteps     = 5,
    refineLR        = 0.01
  ),
  verbose = FALSE
)
```

# Visualization with visualizeHIVE

The `visualizeHIVE()` function produces publication-ready ggplot2 figures
for bHIVE and honeycombHIVE results.

## Plot Types

| Type | Description |
|:-----|:------------|
| `"scatter"` | 2D scatter with optional dimensionality reduction (PCA, UMAP, t-SNE) |
| `"boxplot"` | Feature distributions by group with prototype overlay |
| `"violin"` | Violin plots with prototype markers |
| `"density"` | Density curves with prototype reference lines |

## Scatter Plot with PCA

```{r viz-scatter, fig.width=7, fig.height=5}
X <- as.matrix(iris[, 1:4])

set.seed(42)
res <- honeycombHIVE(X = X, task = "clustering", layers = 3,
                     nAntibodies = 30, beta = 5, epsilon = 0.05,
                     maxIter = 10, verbose = FALSE)

visualizeHIVE(result = res, X = iris[, 1:4],
              plot_type = "scatter",
              transformation_method = "PCA",
              title = "Clustering (Layer 2)",
              layer = 2, task = "clustering")
```

## Violin Plot for Classification

```{r viz-violin, fig.width=7, fig.height=4}
set.seed(42)
res_cls <- honeycombHIVE(X = X, y = iris$Species,
                         task = "classification",
                         layers = 2, nAntibodies = 15,
                         beta = 5, maxIter = 10, verbose = FALSE)

visualizeHIVE(result = res_cls, X = iris[, 1:4],
              plot_type = "violin", feature = "Petal.Length",
              title = "Petal Length by Predicted Class",
              layer = 1, task = "classification")
```

# Combining Modules with honeycombHIVE

The R6 API with modules and the functional `honeycombHIVE()` API serve
different use cases. For maximum control, use `AINet` directly with modules
for single-pass analysis. For hierarchical refinement, use `honeycombHIVE()`
with its built-in layer management and gradient refinement.

A typical advanced workflow:

1. **Explore** with `bHIVE()` to establish baseline performance
2. **Tune** with `swarmbHIVE()` to find good hyperparameters
3. **Compose** modules via `AINet` to add biological mechanisms
4. **Layer** with `honeycombHIVE()` for hierarchical refinement
5. **Visualize** with `visualizeHIVE()` to understand results

```{r}
sessionInfo()
```
